Nengo: A Python tool for building large-scale functional brain models

### 2013-10-02 nengo.tex
Edits from James's feedback

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail
with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological connections \cite{TODO}; % blue brain
cognition has not yet emerged
from data-driven large scale models.
Similarly, theoretical frameworks
that use a generic algorithm
replicated many times % reword based on papers below
are unable to solve more complex problems
than the generic algorithm alone
(e.g., \cite{TODO} \cite{TODO}). % Hawkins, Kurzweil

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the memorized list,
it would trace out the digits $1234$.

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules.
Each functional module corresponds
to a brain area that has been hypothesized
to perform those functions
in neuroscientific literature
(e.g., motor commands are represented
in primary motor cortex).
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is possible due to the NEF,
which, unlike the memory prediction framework
and theoretical frameworks,
allows Spaun to employ
different sets of algorithms
depending on the behavioral task performed,
rather than defining a single algorithm
and hypothesizing that replicating it
many times will enable these behaviors.

In implementing Spaun,
we provided strong support
that the NEF is beginning to fill the theory void in neuroscience.
However, in doing so, we pushed Nengo
to the limits of its ability to construct and simulate
large-scale neural models.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo with a focus on speed,
extensibility, and simplicity.

@@

Finally, perhaps the most significant limitation
of the JavaNengo tools is performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that neural models require quickly and in parallel.

Addressing all of these issues
in the Java codebase would require
such extensive changes
that we instead opted
to write a new version of Nengo in Python.
Since the scripting interface
already used Python syntax,
and since Python is widely used
for neuroscientific data analysis,
we targeted the CPython interpreter
as the platform for the new version of Nengo.

@@

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in applications and games.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
PyNengo's scripting interface to those objects
is modified from JavaNengo's scripting interface \cite{TODO}
to easily incorporate user-defined networks,
new neuron models and learning rules,
and multiple simulators.
Additionally, the scripting interface
contains functions to record data during a simulation,
and access that data after the simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage special-purpose hardware.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

PyNengo does not currently
replace all of JavaNengo's functionality.
While it can simulate the vast majority
of models that JavaNengo can simulate,
some functionality is still in development.
Most notably, there is no graphical interface
to PyNengo, and therefore simulations
cannot be visualized while they are running.
However, we are confident that these tools
can be created, and will be much easier
to implement compared to JavaNengo's GUI.

PyNengo overcomes JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using 11\% as many lines of code,\footnote{
  JavaNengo is 40,000 lines of code, not including the GUI,
  and PyNengo is 5000 lines of code. TODO}
and depends on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large model TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is well tested and
easy to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

@@

\begin{enumerate}
  \item \textbf{Representation:} A population of neurons
    collectively represents a time-varying vector of real numbers.
  \item \textbf{Transformation:} Functions on those vectors
    are computed by the connections between populations.
  \item \textbf{Dynamics:} The vectors represented
    by neural populations can be considered state variables
    in a dynamical system, and can be analyzed using control theory.
\end{enumerate}

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We represent information
with time-varying vectors of real numbers
in order to manipulate that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
We can then estimate
the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how likely
that neuron is to fire for a given input signal.
The tuning curve summarizes the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the encoding weight
(the part of the vector space
that causes the neuron to be the most active).
This results in a predictable pattern of firing
given the input signal.

In the decoding process,
the spike trains are first filtered
with an exponentially decaying filter.
Those filtered spike trains are summed together
with weights that are determined
by solving a least-squares minimization problem.
In the bottom panel of Figure~\ref{fig:nef}B,
the red line is the filtered spiking activity
weighted by decoding weights
that minimize the difference between
the actual encoded vector
and the decoded estimate.
The purple line is the filtered spiking activity
weighted by decoding weights
that minimize the different between
the cosine of the actual encoded vector
and the decoded estimate.
Note that these decoding weights are not determined
based on the input signal;
instead, the minimization occurs
on points that are randomly sampled
in the vector space
that the population represents.

\paragraph{Transformation}
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.

@@

Note that, in practice, we rarely use
the full connection weight matrix,
and instead store
the encoding and decoding weights
(i.e., the factors of the connection weight matrix).
This provides significant
space and time savings.

\paragraph{Dynamics}
While doing feedforward transformations
on vectors is sufficient to model
many neural systems,
many systems, notably memory systems,
require persistent activity through recurrent connections.
When recurrent connections are introduced,
the vectors that populations represent
can be thought of as state variables
in a dynamical system.
Neural systems can therefore
be built and analyzed using
the methods of control theory.

@@

\paragraph{Nengo}
Large models can be built
by using of the principles of the NEF
as connectable components
that describe neural systems,
just as a circuit diagram
describes an electronic circuit.
The goal of Nengo is to enable
modellers to create those components
and connect them.
Components describe
what information is being represented,
and connections describe
how that information is transformed
over time.
Nengo translates those descriptions
to a network of interconnected neurons,
situating it as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model.

@@

\textt{Network} is a base class designed to be
subclassed by users
in order to group together
parts of a model.
In most cases,
the code that creates and connects
several objects in the model can be
made into a Network subclass
with only minor changes.

@@

Nengo is a young project that
has started with a deliberately minimal base
in order to make contributing easier than in
previous implementations.
Additionally, many neural modellers
are already using Python,
as evidenced by the existence of
PyNN, Brian, and Python bindings to other simulators.
Nengo will be able to benefit
from improvements to scientific Python tools.
It is also extendable by modellers with
varying technical skills;
creating a new network is simple,
while creating a new simulator is complex.

By enabling outside contributions,
and providing a user friendly API
and many example models,
we hope to foster a rich community of modellers
contributing networks, nonlinearities,
and simulators to Nengo
and to theoretical neuroscience
as a whole.
Internally, we will bring the functionality
of PyNengo to parity with JavaNengo
by implementing the use cases
not currently covered by PyNengo,
and create a graphical interface
to build models and
interactively inspect simulations.

>>>>

Modeling the brain of a human,
or just about any animal at all,
is one of the greatest
scientific challenges of our time.
However, despite decades of detailed
experimental work in neurobiology
and the recent growth of computational neuroscience,
we lack a mathematical account of
how biological and electrical components
implement cognitive functions such as sensory processing,
memory formation, reasoning, and motor control.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological connections \cite{TODO}. % blue brain
Cognition has not yet emerged
from data-driven large scale models.
On the other end of the spectrum,
cognitive architectures \cite{TODO} % act-r
and theoretical frameworks \cite{TODO} % Hawkins, Kurzweil
have seen success solving statistical problems,
but do so without respecting biological constraints.
Additionally, the statistical problems
solved by these frameworks are better solved
by machine learning approaches.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
The NEF is a large-scale modeling approach
that can leverage any neuron model
to build neural networks with
demonstrable cognitive abilities.
Nengo and the NEF has been used to build
increasingly sophisticated neural subsystems
for the last decade (e.g., subsystem \cite{TODO}),
culminating recently with Spaun,
currently the world's
largest functional brain model \cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the memorized list,
it would trace out the digits $1234$.
While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules.
Each functional module corresponds
to a brain area that has been hypothesized
to perform those functions
in neuroscientific literature.

The NEF shows how to
construct a neural model
from anatomical constraints,
cognitive objectives, and
dynamical systems theory.
Perhaps surprisingly,
constructing models from this starting point,
rather than from single cell electrophysiology
and connectivity statistics
has not meant sacrificing
explanatory power.
Instead, single cell activity,
response timing, and behavioral errors
of NEF-designed models match
psychological and physiological findings
without being built specifically into the design.
These quirks seem to fall directly out
of the need to satisfy competing objectives
within anatomical and neurobiological constraints.
The opposite phenomenon in which
cognitively useful behaviors emerge
from networks of detailed neurons
connected with plausible connectivity statistics
has not yet been observed.

It has often been said that neuroscience
is data-rich and theory-poor.
The modeling success of
large functional networks such as Spaun
indicates that the NEF
is beginning to fill the theory void.
However, in simulating these large networks,
we pushed our Java-powered simulator
to its limits in terms of memory consumption,
simulation speed, and extensibility.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo with a focus on speed,
extensibility, and simplicity.

This paper introduces a new version
of Nengo (PyNengo) written in Python.
The remainder of this introduction outlines
the design issues in the previous version (JavaNengo)
that motivated the reimplementation.
PyNengo is not currently meant
to completely replace JavaNengo;
JavaNengo is a GUI application
and it remains a capable visual NEF
simulator and valuable teaching tool.
PyNengo's focus is on speed
and a simplified internal design
to support ongoing research in
large adaptive functional models.
The remainder of the paper outlines PyNengo's design,
scripting interface, and performance relative to
alternatives such as Brian and NEST.

@@

Finally, perhaps the most significant limitation
of the JavaNengo tools is performance.
While Java has good support for threads,
it lacks native high-performance numerical libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which are well-suited to the computations
required of neural simulator.

Addressing all of these issues
in the Java codebase would require
such extensive changes
that we instead opted
to write a new version of Nengo in Python.
JavaNengo's Jython scripting interface
already used Python syntax,
and since Python is widely used
for neural simulation and data analysis,
we targeted the CPython interpreter
as the platform for the new version of Nengo.

@@

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
Its interfaces are based on NumPy,
so it integrates seamlessly
with other scientific Python tools
for interactive exploration \cite{TODO}, %Ipython
scientific computation \cite{TODO}, %scipy
and plotting \cite{TODO}. % Matlab
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in web services, games, and other applications.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
These objects are accessible
through a scripting interface
similar to JavaNengo's \cite{TODO}, % frontiers 2009
with improvements
for easier incorporation of user-defined networks,
neuron models and learning rules.
Critically, and unlike JavaNengo,
PyNengo's scripting interface
contains functions to record simulated data
and access that data after simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulation packages
(e.g., those accessible by PyNN)
or for faster simulators to be used
as drop-in replacements for PyNengo's
platform-independent reference simulator.
To date, we have implemented one other simulator
that uses PyOpenCL to take advantage
of a GPU or multicore CPU.

PyNengo does not currently
replace all of JavaNengo's functionality.
While it can simulate most
models that JavaNengo can simulate,
there is no graphical interface to PyNengo.
Model components must be created
using Python scripts rather than
the drag-and-drop interface presented by JavaNengo.
Some of the means of interacting
with a running model in JavaNengo
can be provided using packages
such as IPython and Matplotlib,
but such tools have not been our focus thus far.
It is possible to build a graphical program
like JavaNengo's GUI on top of PyNengo's object model,
but that remains future work.

To summarize, PyNengo overcomes
JavaNengo's limitations as exposed by large-scale modelling research.
It can simulate most models
currently simulated by JavaNengo
using a well-tested codebase of
TODO lines of code,
compared to JavaNengo's TODO
(which does not include the JavaNengo GUI).
A new internal representation of neural models
allows the OpenCL simulator to simulate
large models on the scale of Spaun
hundreds of times faster than JavaNengo's simulator
using inexpensive commodity hardware.
Compatibility with CPython makes scientific Python software
available for preparing and analyzing data.
In all, PyNengo provides a platform for
simulating larger and more complex models than Spaun,
and can therefore further test NEF
as a theory of the brain.

@@

\begin{enumerate}
  \item \textbf{Representation:} A population of neurons
    collectively represents a time-varying vector of real numbers.
  \item \textbf{Transformation:} Functions on those vectors
    are computed by the connections between populations of neurons.
  \item \textbf{Dynamics:} The vectors represented
    by neural populations can be considered state variables
    in a dynamical system, and can be analyzed using control theory.
\end{enumerate}

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We represent information
with time-varying vectors of real numbers
in order to manipulate that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
With enough neurons,
we can estimate the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how likely
that neuron is to respond to a given input signal.
The tuning curve is a function of the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the \textit{encoding weight}
(the direction in the input vector space
that causes the neuron to be the most active).
% This results in a predictable pattern of firing
% given the input signal.

In the decoding process,
the spike trains are first filtered
with an exponentially decaying filter.
Those filtered spike trains are summed together
with weights that are determined
by solving a least-squares minimization problem.
In the bottom panel of Figure~\ref{fig:nef}B,
the red line is the filtered spiking activity
weighted by decoding weights
that minimize the difference between
the actual encoded vector
and the decoded estimate.
The purple line is the filtered spiking activity
weighted by decoding weights
that minimize the different between
the cosine of the actual encoded vector
and the decoded estimate.
Note that these decoding weights
need not be determined
based on the input signal;
instead, we typically perform
this minimization on points
randomly sampled from
the portion of the vector space
that the population represents.

\paragraph{Transformation}  % NEF as neural model...
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.
In other words, the NEF hypothesizes
that synaptic weight matrices
have low rank, and can be factored.

@@

Note that, in practice, we rarely use
the full connection weight matrix,
and instead store
the encoding and decoding weights separately
(i.e., the two factors of the connection weight matrix).
This provides significant
space and time savings in the simulation loop.

\paragraph{Dynamics}
While feedforward vector transformations
suffice to describe
some neural systems,
many systems, notably memory systems,
require persistent activity through recurrent connections.
When recurrent connections are introduced,
the vectors represented by neural populations
can be thought of as state variables
in a dynamical system.
The equations governing dynamics
in such a system
can be designed and analyzed
using the methods of control theory,
and translated into neural circuitry
using the principles
of representation and transformation.

@@

\paragraph{NEF and Nengo}
% The NEF makes it possible
% to design large systems
% because its three principles
% show how to compose functional subsystems
% into a coherent whole.
Large models can be built
by using of the principles of the NEF
as connectable components
that describe neural systems,
just as a circuit diagram
describes an electronic circuit.
The goal of Nengo is to enable
modellers to create those components
and connect them.
Components describe
what information is being represented,
and connections describe
how that information is transformed
over time.
Nengo translates those descriptions
to a network of interconnected neurons,
situating it as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model.

@@

\texttt{Network} is a base class designed to be
subclassed by users
in order to group together
parts of a model.
In most cases,
the code that creates and connects
several objects in the model can be
made into a Network subclass
with only minor changes.

@@

Nengo is a young project that
has started with a deliberately minimal base
in order to make contributing easier than in
previous implementations.
Additionally, many neural modellers
are already using Python,
as evidenced by the existence of
PyNN, Brian, and Python bindings to other simulators.
Nengo will be able to benefit
from improvements to scientific Python tools.
It is also extendable by modellers with
varying technical skills;
creating a new network is simple,
while creating a new simulator is complex, but possible.

By enabling outside contributions,
and providing a user-friendly API
and many example models,
we hope to foster a rich community of modellers
contributing networks, nonlinearities,
and simulators to Nengo
and to theoretical neuroscience
as a whole.
Internally, we will bring the functionality
of PyNengo to parity with JavaNengo
by implementing the use cases
not currently covered by PyNengo,
and create a graphical interface
to build models and
interactively inspect simulations.
